Work Fragmentation in Developer Interaction Data

### 2016-07-04 paper_2.tex
typos fixed

Work fragmentation is indeed an important phenomenon within the context of modern software development, and the impact there may be even worse since developers build and maintain complex mental models of the software they are working on---these models may be more sensible to interruptions, and costly to rebuild. %; and 2) development is more often distributed, which may vary the nature of interruptions (trading high-bandwidth meetings for low-bandwidth electronic messages). Achieving flow in these conditions may be challenging.

In this paper, we present a study addressing these issues by (1) being focused on software developers, and (2) using MSR (Mining Software Repositories) techniques to base its conclusions on a much larger amount of data (specifically, thousands of development sessions recorded by Mylyn and the Eclipse Usage Data Collector). The downside is that the data exploited, being operational data \cite{M14}, is of lower contextual quality than data extracted from a full-fledged observational study. Trading quality for quantity allows us to explore different aspects of the phenomenon, and to generate hypotheses for subsequent studies of work fragmentation and interruptions in software development. We address the following three questions:
\begin{itemize}
\item \textbf{RQ1:} What is the relationship between the observed interruptions and the observed developer productivity? Every interruption requires time to regain the lost context, so in this research question we intend to see whether there is an accumulative effect when the number of interruption increases. In this context, we define productivity as the amount of activity within the IDE. 
\item \textbf{RQ2:} Is the observed relationship more pronounced in the presence of prolonged interruptions? A longer interruption might be more detrimental to the productivity than several small interruptions, as it requires more time to recover from it; this research question explores that possibility.
\item \textbf{RQ3:}  What is the observed relationship in the vicinity near of interruptions? Based on the literature and observational studies, we believe that in the immediate time around an interruption the effect of it can be more palpable.
\end{itemize}

@@

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). A MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon and it is a way of transforming a spatial-temporal model to a simple time serie model, based on the work by Rohrdantz et al. \cite{RHD12}.

@@

In the results with Mylyn, we observe a large difference between the sessions without interruptions and the ones who do. Further, we observe that the rate of events per minute decreases slightly when the session has more interruptions. Therefore, we can intuit that the relationship between number of interruptions and the productivity indicators that are edits and selections, tends to be inversely proportional.

@@

More importantly, we used \textit{Cliff's delta} to measure the practical significance of these results in term of effect size % the effect size of the differences in means as standardized difference~
\cite{C94}. \textit{Cliff's delta thresholds} are defined as follows: negligible ($<0.147$), small  ($<0.33$), moderate ($<0.476$) and large effect otherwise. As shown in Table \ref{tbl:p_value}, we note that the effect size of the interruptions over the number of edits by minute is large. In selections, the effect is moderate for sessions having up three interruptions, and large for sessions with over four interruptions. This reinforces our impressions that interruptions and user activity follow inverse relationships, and that they are quite pronounced.

@@

We first analyze the relationship between edit ratio and number of interruptions in Mylyn (Figure \ref{fig:box_int_events}). We observe that the edit ratio decreases when the session has more interruptions: the effect is pronounced between session that do not have interruptions and ones that do, and is more subtle as the number of interruption grows. A look at the practical and statistical significance of these results (Table \ref{tbl:p_value}, bottom) show that the results are (unsurprisingly) statistically significant, and that the observed effect sizes are large to very large.

@@

The Figure \ref{fig:dur_mylyn} shows the median of the metrics when grouping the sessions by the number of interruptions and the proportion of prolonged interruptions with Mylyn, and the Figure \ref{fig:dur_udc} shows the same with UDC. According to our hypothesis, we expect sessions that have a low number of interruptions and a low proportion of prolonged interruptions to have higher rates of productivity, and the results mostly agree with this statement. First, we observe that (as we conclude in RQ1) when the number of interruption increases the median of the metric decreases until reaching a point (approximately between three and four interruptions) where the change is minimal and continues that way for the rest of the groups. Second, in the vast majority of groups the median value of the \emph{low} group is greater than the \emph{high} group---this happens 35 times, while the opposite is true twice, and the median are equals five times. This is the clearest in the edits per minute metric in both datasets, as we can see that at some points the medians intertwine in the selections per minute and edit ratio, particularly in Mylyn.

@@

Since these results apply to both datasets, we can conclude that we observe a relationship between the duration of the interruptions and the productivity, with a caveat: the relation is much more pronounced when the amount of interruptions is lower; as the number of interruptions grows it becomes very subtle. We speculate, but can not confirm, that this is due to the noise in the dataset, as sessions with more interruptions are usually longer and thus may exhibit more variation due to other factors.

@@

In general, the effect of positive and negative interruptions is seems to be reversed. The editions per minute describes better the effect and the selections do not show a major change. There is not a clear pattern during a neutral interruption.

>>>>

Work fragmentation is indeed an important phenomenon within the context of modern software development, and the impact there may be even worse since developers build and maintain complex mental models of the software they are working on---these models may be more sensitive to interruptions, and costly to rebuild. %; and 2) development is more often distributed, which may vary the nature of interruptions (trading high-bandwidth meetings for low-bandwidth electronic messages). Achieving flow in these conditions may be challenging.

In this paper, we present a study addressing these issues by (1) being focused on software developers, and (2) using MSR (Mining Software Repositories) techniques to base its conclusions on a much larger amount of data (specifically, thousands of development sessions recorded by Mylyn and the Eclipse Usage Data Collector). The downside is that the data exploited, being operational data \cite{M14}, is of lower contextual quality than data extracted from a full-fledged observational study. Trading quality for quantity allows us to explore different aspects of the phenomenon, and to generate hypotheses for subsequent studies of work fragmentation and interruptions in software development. We address the following three questions:
\begin{itemize}
\item \textbf{RQ1:} What is the relationship between the observed interruptions and the observed developer productivity? Every interruption requires time to regain the lost context, so in this research question we intend to see whether there is an accumulative effect when the number of interruption increases. In this context, we define productivity as the amount of activity within the IDE. 
\item \textbf{RQ2:} Is the observed relationship more pronounced in the presence of prolonged interruptions? A longer interruption might be more detrimental to the productivity than several small interruptions, as it requires more time to recover from it; this research question explores that possibility.
\item \textbf{RQ3:}  What is the observed relationship in the vicinity of interruptions? Based on the literature and observational studies, we believe that in the immediate time around an interruption the effect can be more evident.
\end{itemize}

@@

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). A MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon and it is a way of transforming a spatial-temporal model to a simple time series model, based on the work by Rohrdantz et al. \cite{RHD12}.

@@

In the results with Mylyn, we observe a large difference between the sessions without interruptions and the ones who have. Further, we observe that the rate of events per minute decreases slightly when the session has more interruptions. Therefore, we can intuit that the relationship between number of interruptions and the productivity indicators that are edits and selections, tends to be inversely proportional.

@@

More importantly, we used \textit{Cliff's delta} to measure the practical significance of these results in term of effect size % the effect size of the differences in means as standardized difference~
\cite{C94}. \textit{Cliff's delta thresholds} are defined as follows: negligible ($<0.147$), small  ($<0.33$), moderate ($<0.476$) and large effect otherwise. As shown in Table \ref{tbl:p_value}, we note that the effect size of the interruptions over the number of edits per minute is large. In selections, the effect is moderate for sessions having up three interruptions, and large for sessions with over four interruptions. This reinforces our impressions that interruptions and user activity follow inverse relationships, and that they are quite pronounced.

@@

We first analyze the relationship between edit ratio and number of interruptions in Mylyn (Figure \ref{fig:box_int_events}). We observe that the edit ratio decreases when the session has more interruptions: the effect is pronounced between session that do not have interruptions and ones that do, and is more subtle when the number of interruptions increases. The Table \ref{tbl:p_value} (bottom) shows that the results are (unsurprisingly) statistically significant, and that the observed effect sizes are large to very large.

@@

The Figure \ref{fig:dur_mylyn} shows the median of the metrics when grouping the sessions by the number of interruptions and the proportion of prolonged interruptions with Mylyn, and the Figure \ref{fig:dur_udc} shows the same with UDC. According to our hypothesis, we expect sessions that have a low number of interruptions and a low proportion of prolonged interruptions to have higher rates of productivity, and the results mostly agree with this statement. First, we observe that (as we conclude in RQ1) when the number of interruptions increases the median of the metric decreases until reaching a point (approximately between three and four interruptions) where the change is minimal and continues that way for the rest of the groups. Second, in the vast majority of groups the median value of the \emph{low} group is greater than the \emph{high} group---this happens 35 times, while the opposite is true twice, and the median are equals five times. This is the clearest in the edits per minute metric in both datasets, as we can see that at some points the medians intertwine in the selections per minute and edit ratio, particularly in Mylyn.

@@

Since these results apply to both datasets, we can conclude that we observe a relationship between the duration of interruptions and the productivity metrics, with a caveat: the relation is much more pronounced when the amount of interruptions is lower; as the number of interruptions grows it becomes very subtle. We speculate, but can not confirm, that this is due to the noise in the dataset, as sessions with more interruptions are usually longer and thus may exhibit more variation due to other factors.

@@

In general, the effect of positive and negative interruptions seems to be reversed. The editions per minute describe better the effect and the selections do not show a major change. There is not a clear pattern during a neutral interruption.
